Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-05-26 docs/parasite_mortality_manuscript.tex
Finished added in Cherie's comments on the main text

Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
require binning or data alteration, reduces the number of
parameters to be estimated, is highly generalizable, and uses standard statistical techniques to determine PIHM significance.
%We provide Python code for implementing the Likelihood Method in \emph{SI} 4.
The Likelihood Method first assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \phi)$, which specifies the probability of a host having $x$ parasites when it is observed.  $\phi$ is a vector of parameters that described the shape and scale of this distribution.

The Likelihood Method then assumes that the probability of a host surviving for
$t$ units given it has some parasite intensity $x$ is described by the function
$h(\text{survival} ; x, t, \theta)$ where $\theta$ is a vector of parameters of
the survival function. For most parasite-intensity datasets, the observer has
no knowledge of how long a host has been infected with $x$.  One way to account
for this would be to integrate out time such that we are left with the function
$h(\text{survival} ; x, \theta)$ specifying the probability of survival given
$x$ parasites and some additional parameters $\theta$.  While eliminating the
functions dependence on time is far from ideal, this is an inherent limitation
of attempting inference on cross-sectional data (jargon).

@@

One can see that $P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \theta)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \phi)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \theta)  * g(x; \phi)$. Therefore equation \ref{eq:concept} can be written as

\begin{equation}
    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \theta)  * g(x; \phi)}{\sum_{x=0}^{\infty} h(\text{survival}; x, \theta)  * g(x; \phi)}
    \label{eq:dist}
\end{equation}

@@

Using this probability distribution, one can then find the parameters $\theta$
and $\phi$ that maximize the likelihood of an observed host-parasite dataset.
To estimate the significance of PIHM in a host-parasite system, a likelihood
ratio test can be used in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by the pre-mortality distribution
$g(x; \phi)$.  If PIHM is not significant in the system, the resulting
likelihood ratio statistic should approximately follow a $\chi^2$ distribution
with two degrees of freedom.

Equation \ref{eq:dist} could be parameterized in many different ways depending
on the parasite system of interest. In this study, we will follow the precedent
set by all previous methods for estimating PIHM and assume that the pre-
mortality parasite distribution $g(x; \phi)$ follows a negative binomial
distribution with the parameters mean parasite intensity ($\mu_p$) and
aggregation ($k_p$) before mortality, respectively (smaller $k_p$ indicates
more aggregation).   The negative binomial distribution is not just a
phenomenological assumption and can arise as the equilibrium parasite
distribution under a variety of different biological assumptions
\citep{Calabrese2011} [MORE]. However, it is also an incredibly flexible
distribution that fits many host-parasite systems regardless of whether the
underlying mechanisms lead to an exact negative binomial distribution
\citep{Shaw1998}.

Choosing a function for $h(\text{survival}; x, \theta)$

 The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.  All parameters are defined in Table \ref{tab:params}.

@@

We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters. First, we created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$. Second, we chose values of $a$ and $b$ and calculated the probability of survival
for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, for each host, we drew a random number from a uniform distribution
between 0 and 1 and if the calculated host survival probability was less than this random
number, the host experienced parasite-induced mortality. The parasite distribution in these simulated surviving hosts is equivalent to the observed parasite distribution in a wild host population that has undergone parasite-induced host mortality.

@@

We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:question1}A).  For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.  We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500]. $N_p$ is not technically the sample size on which the methods are being
tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.\\

\noindent
\emph{Question 2: Can we estimate fatal parasite intensity and the host survival function?}

To compare the ability of the Adjei Method and the Likelihood Method to recover the $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above. For each parameter
combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates over pre-mortality host population sizes of $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]. We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.  Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.

\subsection*{Efficacy of the Likelihood Method with unknown pre-mortality parameters}

In the final simulation, we tested the ability of the Likelihood Method to
correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality
parameters are unknown. The previous simulations showed that the Likelihood
Method effectively identified PIHM when $\mu_p$ and $k_p$ were known with
values of 10 and 1, respectively.  As a best-case scenario, we simulated host-
parasite systems with these pre-mortality parameters and tested the power of
the Likelihood Method to identify PIHM for gradual, moderate and steep survival
functions when the pre-mortality parameters $\mu_p$ and $k_p$ also needed to be estimated.  We
perform 500 simulations over a range of different samples sizes following the
simulation procedure described above.

@@

When all parameters were jointly estimated, the Likelihood Method showed highly
context-dependent results when detecting PIHM in the best-case scenario of
$\mu_p = 10$ and $k_p = 1$. The Likelihood Method's power of detecting PIHM was
greater than 0.8 when when host sample sizes were 424 and 83  for survival
functions that were moderate and steep, respectively (Fig
\ref{fig:real_power}).  When the host survival function was gradual, the
Likelihood Method never had a power greater than 0.8 for any post-mortality
samples sizes we considered (Fig \ref{fig:real_power}).

@@

Of the 10 datasets we considered, the previous authors qualitatively detected PIHM
in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
from the pre-mortality parameters of the Crofton Method detected significant
PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
dataset in which the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
for female \emph{S. tumbil}.  For this dataset there was a marginally significant effect
of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$). The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results that the Adjei Method has a high Type I error rate.

@@

Quantifying the impact of parasitism on wild host populations is critical for managing wildlife populations and understanding parasite transmission. Ideally the relationship between
infection intensity and host survival would be measured experimentally, but for
logistical and ethical reasons, this is often impossible \citep{McCallum2000a}.
Looking for evidence of mortality in parasite distribution data requires the
least amount of information, but is notoriously difficult to implement. The
methodological flaws in the Adjei Method limit its utility, so here we propose
an alternative, likelihood-based, method to estimate host survival and the
$LD_{50}$ from observed parasite intensity data.  This
method is a significant improvement over the previous methods because it requires fewer parameters,
provides a statistical decision rule for identifying PIHM and does not require
any data manipulation.

Using simulated data, we found that the Likelihood Method always out performed the Adjei Method. For simply detecting the presence of PIHM, the Likelihood
Method was both more powerful and had fewer false detection events (Type I
errors).  When both methods were applied to published datasets previously used
in PIHM analyses, the Adjei Method tended to detect PIHM where it had not previously been
reported, consistent with the high Type I error rate observed in our
simulations. The Likelihood Method was also more precise and less
biased in calculations of both the parasite $LD_{50}$ and host survival curve over the parameter values we considered.
However, while only the Likelihood Method produced precise and unbiased $LD_{50}$
estimates, neither method could provide unbiased estimates of the host survival
function at realistic sample sizes.  These simulations demonstrate that
the Likelihood Method is more powerful and precise than the previously propose Adjei Method.

Although superior to the Adjei Method, the Likelihood Method is not universally applicable to real data.  Our simulations showed when the when pre-
mortality parameters were estimated directly, the Likelihood Method needed at
least 83-424 samples to have 80\% power and for steep to moderate survival functions and even
more as the survival function became more gradual. While some of these sample sizes are reasonable for hosts such as invertebrates or small fish, even the smallest sample sizes are completely
unfeasible for many vertebrates, particularly the species of conservation
concern where addressing the impact of parasitism would be most important. An
even larger sample size would be required to identify PIHM when parasites are highly aggregated, mean infection intensity is
high, or parasite prevalence is low, all of which are
common in many parasitic helminths.  Moreover, our results are in agreement with previous work that has shown that as host-survival functions become progressively more linear, PIHM becomes all but impossible to detect \citep{Lanciani1989}.  This result, however, does not preclude the use of this method as non-linear survival functions
are not uncommon in empirical host-parasite systems \citep{Benesh2011}.  Finally,
while linear functions make PIHM undetectable, at the other extreme, steep,
non-linear survival curves produce severely biased estimates of the survival
function. Give the interaction between all of these different factors, the
Likelihood Method is probably limited to detecting PIHM in systems where greater than 100 hosts can be collected, parasites are
common and only moderately aggregated, and substantial host mortality occurs at relatively low parasite intensity.

@@

Given these numerous caveats, is there a place in parasitology for methods that
estimate PIHM from intensity data?  We are in agreement with
\cite{Lester1984} that, at the very least, methods for estimating PIHM can
provide preliminary insight into whether or not PIHM is worth further
exploration.  However, we stress that these methods should only be used as an
exploratory tool when assessing the role of PIHM in a system and potential
users should critically evaluate whether they think they have a large enough
sample size and an appropriate host survival function/post-mortality distribution for the methods developed
in this paper to be applicable.  Even if they are applicable, inferring PIHM
from distributional data is no substitute for field experiments
and an in depth understanding of the natural history of the host-parasite
system under consideration.

@@

\begin{table}
    \centering
    \caption{Definition of parameters and functions used in the main text}
    \begin{tabular}{c c}
    \hline
    Parameter & Definition \\
    \hline\hline
    $\mu_p$ & Pre-mortality mean parasite intensity \\
    $k_p$   & Pre-mortality parasite aggregation \\
    $N_p$   & Pre-mortality host population size \\
    $x$     & Number of parasites in a given host \\
    $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial parasite distribution \\
    $a$ & Parameter of the logistic host survival function \\
    $b$ & Parameter of the logistic host survival function \\
    $h(x; a, b)$ & The host survival function \\
    $LD_{50}$ & $\exp(a / b)$, parasite intensity at which 50\% of hosts die \\

>>>>

Here we propose a novel, likelihood-based method (henceforth Likelihood Method)
that does not require binning or data alteration, reduces the number of
parameters to be estimated, is highly generalizable, and uses standard
statistical techniques to determine PIHM significance.  The Likelihood Method
begins with the same assumptions as the Adjei Method: namely that infection of
a host, parasite-induced mortality of a host, an the sampling of a host
population occur at distinct time intervals during a hosts life. As discussed by
\citeauthor{Adjei1986}, this is not necessarily unrealistic as studies have
shown that infection and host mortality are often age and/or body size dependent [citations].

The Likelihood Method then assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \boldsymbol{\phi})$, which specifies the probability of a host having $x$ parasites when it is observed.  $\boldsymbol{\phi}$ is a vector of parameters that describes the shape this distribution.

The method then assumes that the probability of a host surviving with $x$ parasites from infection until sampling is given by $h(\text{survival} ; x, \boldsymbol{\theta})$ where $\boldsymbol{\theta}$ specifies any additional parameters needed to define the host survival function.

@@

$P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \boldsymbol{\theta})$, $P(x)$ is the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \phi)$. Therefore equation \ref{eq:concept} can be written as

\begin{equation}
    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}{\sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}
    \label{eq:dist}
\end{equation}

@@

Using this probability distribution, one can then find the parameters $\boldsymbol{\theta}$
and $\boldsymbol{\phi}$ that maximize the likelihood of an observed host-parasite dataset.
To estimate the significance of PIHM in a host-parasite system, a likelihood
ratio test can be used in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by the pre-mortality distribution
$g(x; \boldsymbol{\phi})$.  If PIHM is not significant in the system, the resulting
likelihood ratio statistic should approximately follow a $\chi^2$ distribution
with degree of freedom equal to the number of parameters in the full model with parasite-induced mortality minus the number of parameters in the reduced model without parasite-induced mortality [citation].

Equation \ref{eq:dist} could be parameterized in many different ways depending
on the parasite system of interest. In this study, we adopt the typical assumption that the pre-
mortality parasite distribution $g(x; \boldsymbol{\phi})$ follows a negative binomial
distribution with the parameters mean parasite intensity ($\mu_p$) and
aggregation ($k_p$) before mortality, respectively (smaller $k_p$ indicates
more aggregation) \citep{Crofton1971a,AndersonandMay1978,Adjei1986}.   The negative binomial distribution can arise as the equilibrium parasite
distribution under a variety of different biological and statistical assumptions
\citep{Kendall1948a, Boswell1970, Calabrese2011}. However, it is also an incredibly flexible
distribution that fits many host-parasite systems regardless of whether the
underlying mechanisms lead to an exact negative binomial distribution
\citep{Shaw1998}.

Choosing an appropriate function for $h(\text{survival}; x, \boldsymbol{\theta})$ will depend on the system under consideration.  Many theoretical models of
parasite-induced host mortality assume that the parasite-induced death rate of
hosts is a linear function of parasite intensity
\citep{AndersonandMay1978,Dobson1992,Barbour2000}. It has been previously noted that parasite-induced mortality can be nearly impossible to detect from intensity data when the host survival function is a linear function of parasite intensity as the post-mortality distribution will be of a similar form as the pre-mortality distribution \citep{Lanciani1989}. For example, \cite{Barbour2000} showed that both simple host parasite models without parasite-induced host mortality and those with linear parasite-induced host mortality produced Poisson distributions at equilibrium. More complex, non-linear survival functions were needed to produce over-dispersed of under-dispersed post-mortality distributions.  This is not unrealistic as empirical evidence has shown that non-linear host-survival functions are not uncommon
\citep{Benesh2011} [phrasing].

As one of the goals of this study is to compare this new
Likelihood Method to the previously proposed Adjei Method, we adopt the host-
survival function used in their study and assume host-survival is non-linear and follows a logistic function given by

\begin{equation}
    h(\text{survival}; x, a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}
where $b / 4$ determines the maximum rate of decline of
host survival probability with increasing parasite load, analogous to the pathogenicity parameter $\alpha$ in traditional macroparasite models \citep{AndersonandMay1978}.  When $b$ is held constant, for every one unit increase in $a$ the parasite intensity at which 99\% of hosts survive increases by $ 1/ b$.  The equation $\exp(a
/ b)$ can also be used to calculate the parasite $LD_{50}$, here defined as the
infection intensity at a host has a 50\% dying.  This function is commonly used in toxicology and survival analysis and has the useful properties of being bounded between 0 and 1 and being differentiable for all $x$.  That being said, it is phenomenological and there is little justification to use it rather than it tends to survival data. However, given that a goal of these analyses is to
compare this method's results the those given by the Adjei Method it is natural
to adopt the same host-survival function to facilitate comparison.  When
applying the likelihood method to other systems, other more mechanistic host-survival functions can be used in place of equation \ref{eq:logistic}.

@@

We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters. Consistent with the assumptions of the model that parasite infection, mortality, and sampling occur at distinct life stages of the host, we first created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$. This is equivalent to the period of hosts becoming infected with parasites given in \cite{Adjei1986}.  In the Adjei Method and Crofton Method, $N_p$ is a necessary parameter that is defined as the number of hosts in the population before parasite-induced mortality.  A more appropriate way to define this parameter is the number of hosts that would have been sampled had parasite-induced host mortality not occurred.  This parameter is not necessary using the Likelihood Method because unlike the Adjei Method and Crofton Method which estimate parasite-induced mortality using absolute numbers of hosts, the Likelihood Method estimates parasite-induced mortality using probabilities. However, to compare the results of the Likelihood Method with the Adjei Method, we specified a value for $N_p$ for all simulations.

Second, we chose values of $a$ and $b$ for the host survival function and calculated the probability of survival
for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, for each host, we drew a random number from a uniform distribution
between 0 and 1 and if the calculated host survival probability was less than this random
number, the host experienced parasite-induced mortality.  This was the period in which hosts died due to infection. The parasite distribution in these simulated surviving hosts represented the parasite distribution in a wild host population that has undergone parasite-induced host mortality.

@@

We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had gradual, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:question1}A).  For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.  We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500]. $N_p$ is not technically the sample size on which the methods are being
tested for the post-mortality data because PIHM reduces $N_p$ for each simulated
dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.\\

In the next simulation, we tested the ability of only the Likelihood Method to
correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality
parameters are unknown.  As a best-case scenario, we simulated host-
parasite systems with $\mu_p = 10$ and $k = 1, 0.5, \text{and} 0.1$ [re-run for 0.5 and 0.1] as it is easier to detect PIHM from small samples sizes when mean parasite intensity is low. We then used
the Likelihood Method to identify PIHM for gradual, moderate and steep survival
functions when the pre-mortality parameters $\mu_p$ and $k_p$ also needed to be estimated.  We
perform 500 simulations over a range of different samples sizes following the
simulation procedure described above.

\noindent
\emph{Question 2: Can we estimate properties of the host survival function?}

In the previous section we compared the ability of the Adjei Method and the Likelihood Method to correctly identify whether or not PIHM was a occurring in a system (i.e. a yes or no answer). In this section we compare the ability of the Adjei Method and the Likelihood Method to estimate properties of the survival function such as the parameters $a$, $b$ and $LD_{50}$.  To do this, we used the same simulation procedure and parameter combinations described above. For each parameter
combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates.  Because estimating properties of the host survival function requires more information than simply detecting PIHM, we used larger values of $N_p$ for this simulation ($N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]).  We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.  Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.

@@

When all parameters were jointly estimated, the Likelihood Method showed highly
context-dependent results when detecting PIHM in the best-case scenario of
$\mu_p = 10$ and $k_p = 1$. The Likelihood Method's power of detecting PIHM was
greater than 0.8 when host sample sizes exceeded 424 and 83  for survival
functions that were moderate and steep, respectively (Fig
\ref{fig:real_power}).  When the host survival function was gradual, the
Likelihood Method never had a power greater than 0.8 for any post-mortality
samples sizes we considered (Fig \ref{fig:real_power}).

@@

The previous authors qualitatively detected PIHM
in 7 of the 10 datasets considered(Table \ref{table:pihm}).  The Likelihood Method parameterized
from the pre-mortality parameters of the Crofton Method detected significant
PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
dataset in which the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
for female \emph{S. tumbil}.  For this dataset there was a marginally significant effect
of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$). The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results that the Adjei Method has a high Type I error rate.

@@

Quantifying the impact of parasitism on wild host populations is critical for managing wildlife populations and understanding parasite-host dynamics. Ideally the relationship between
infection intensity and host survival would be measured experimentally, but for
logistical and ethical reasons, this is often impossible \citep{McCallum2000a}.
Looking for evidence of mortality in parasite distribution data requires the
least amount of information, but is notoriously difficult to implement. The
methodological flaws in the Adjei Method limit its utility, so here we propose
an alternative, likelihood-based, method to estimate host survival and the
$LD_{50}$ from observed parasite intensity data.  This
method is a significant improvement over the previous methods because it requires fewer parameters,
provides a statistical decision rule for identifying PIHM and does not require
any data manipulation.

Using simulated data, we found that the Likelihood Method always out performed the Adjei Method. For simply detecting the presence of PIHM, the Likelihood
Method was both more powerful and had fewer false detection events (Type I
errors).  When both methods were applied to published datasets previously used
in PIHM analyses, the Adjei Method tended to detect PIHM where it had not previously been
reported, consistent with the high Type I error rate observed in our
simulations. The Likelihood Method was also more precise and less
biased in calculations of both the parasite $LD_{50}$ and host survival curve over the parameter values we considered.
However, while only the Likelihood Method produced precise and unbiased $LD_{50}$
estimates, neither method could provide unbiased estimates of the host survival
function at realistic sample sizes.  These simulations demonstrate that
the Likelihood Method is more powerful and precise than the previously proposed Adjei Method.

Although superior to the Adjei Method, the Likelihood Method is not universally applicable to real data.  Our simulations showed when the when pre-
mortality parameters were estimated directly, the Likelihood Method needed at
least 83-424 samples to have 80\% power for steep to moderate survival functions, an even larger sample size as the survival function became more gradual. While some of these sample sizes are reasonable for hosts such as invertebrates or small fish, even the smallest sample sizes are completely
unfeasible for many vertebrates, particularly the species of conservation
concern where addressing the impact of parasitism would be most important. An
even larger sample size would be required to identify PIHM when parasites are highly aggregated, mean infection intensity is
high, or parasite prevalence is low, all of which are
common in many parasitic helminths.  Moreover, our results are in agreement with previous work that has shown that as host-survival functions become progressively more linear, PIHM becomes all but impossible to detect \citep{Lanciani1989}.  This result, however, does not preclude the use of this method as non-linear survival functions
are not uncommon in empirical host-parasite systems \citep{Benesh2011}.  Finally,
while linear functions make PIHM undetectable, at the other extreme, steep,
non-linear survival curves produce severely biased estimates of the survival
function. Give the interaction between all of these different factors, the
Likelihood Method is probably limited to detecting PIHM in systems where greater than 100 hosts can be collected, parasites are
common and only moderately aggregated, and substantial host mortality occurs at relatively low parasite intensity.

@@

Given these numerous caveats, is there a place in parasitology for methods that
estimate PIHM from intensity data?  We are in agreement with
\cite{Lester1984} that, at the very least, methods for estimating PIHM can
provide preliminary insight into whether or not PIHM is worth further
exploration.  However, we stress that these methods should only be used as an
exploratory tool when assessing the role of PIHM in a system, and potential
users should critically evaluate whether they think they have a large enough
sample size and an appropriate host survival function/post-mortality distribution for the methods developed
in this paper to be applicable.  Even if they are applicable, inferring PIHM
from distributional data is no substitute for field experiments
and an in depth understanding of the natural history of the host-parasite
system under consideration.

@@

\begin{table}
    \centering
    \caption{Definition of parameters and functions used in the main text}
    \begin{tabular}{c p{12cm}}
    \hline
    Parameter & Definition \\
    \hline\hline
    $\mu_p$ & Pre-mortality mean parasite intensity \\
    $k_p$   & Pre-mortality parasite aggregation \\
    $N_p$   & Pre-mortality host population size \\
    $x$     & Number of parasites in a given host \\
    $g(x; \mu_p, k_p)$ & Pre-mortality negative binomial parasite distribution \\

    $h(\text{survival}; x, a, b)$ & The probability of host survival given a parasite load $x$ and logistic parameters $a$ and $b$\\

    $b / 4$ & The maximum rate of decline in host survival probability with increasing parasite load  \\
    $a$ & When $b$ is held constant a one unit increase in $a$ leads to a $1 / b$ increase in the parasite intensity at which 99\% of hosts survive\\
    $LD_{50}$ & $\exp(a / b)$, parasite intensity at which a host has a 50\% chance of dying \\
